---
title: "Assignment_1"
author: "Stephen Gombos"
date: "2025-09-07"
output:
  word_document: default
  pdf_document: default
---

# 1.) Download a dataset from the web. You may use any source, but specify the source in your code. Also ensure that the data has a mix of quantitative and qualitative (categorical) variables.**

##Done outside of RStudio

## 1.1) Set up RStudio file

```{r}
##install.packages("dplyr") Commented out, only need to install once
library(dplyr)
```

# 2.) Import the dataset into R

```{r}
BathSoap <- read.csv("/Users/stephengombos/Documents/KSU MBA PROGRAM/Fall 2025 FUNDAMENTALS OF MACHINE LEARNING (BA-64060-002)/CSV Files/BathSoap.CSV")
```

## 2.1) Visualize the Dataset

```{r}
head(BathSoap)
```

# 3) Print out descriptive statistics for a selection of quantitative and categorical variables.

```{r}
summary(BathSoap)
```

# 4) Transform at least one variable. It doesn't matter what the transformation is.

First lets look at the column names

```{r}
colnames(BathSoap)
```

Next lets confirm the dataclass for "Pur.Vol.No.Promo...."

```{r}
class(BathSoap$Pur.Vol.No.Promo....)
```

The data in this column is a percentage. For the transfirmation we will remove the % sign and transform it to a number and then to a decimal

```{r}
BathSoap$Pur.Vol.No.Promo.... <- gsub("%", "", BathSoap$Pur.Vol.No.Promo....)
BathSoap$Pur.Vol.No.Promo.... <- as.numeric(BathSoap$Pur.Vol.No.Promo....)
BathSoap$Pur.Vol.No.Promo.... <- BathSoap$Pur.Vol.No.Promo.... / 100
```

confirm the class has changed

```{r}
class(BathSoap$Pur.Vol.No.Promo....)
```

verify data

```{r}
head(BathSoap$Pur.Vol.No.Promo....)
```

# 5) Plot at least one quantitative variable, and one scatterplot

## 5.1) Quantitative variable plot of the newly transformed BathSoap\$Pur.Vol.No.Promo.... using a histogram

```{r}
hist(BathSoap$Pur.Vol.No.Promo....)
```

## 5.2) Scatterplot of the Total Volume vs Value

```{r}
plot(BathSoap$Total.Volume, BathSoap$Value,
     main = "Total Volume vs. Value",
     xlab = "Total Volume",
     ylab = "Value",
     col = "blue")
```
