---
title: "Assignment4(3)_StephenGombos"
author: "Stephen Gombos"
date: "2025-10-25"
output: word_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r}
library(tidyverse)  # For data manipulation and visualization
library(factoextra) # For clustering algorithms & visualization
library(flexclust)    # For k-medians clustering
library(knitr)      # For pretty-printing tables
library(dplyr)      # Specifically for data wrangling
```

## Purpose

The purpose of this assignment is to use k-Means for clustering.

## Direction

An equities analyst is studying the pharmaceutical industry and would
like your help in exploring and understanding the financial data
collected by her firm. Her main objective is to understand the structure
of the pharmaceutical industry using some basic financial measures.
Financial data gathered on 21 firms in the pharmaceutical industry are
available in the file Pharmaceuticals.csv. For each firm, the following
variables are recorded: 1. Market capitalization (in billions of
dollars) 2. Beta 3. Price/earnings ratio 4. Return on equity 5. Return
on assets 6. Asset turnover 7. Leverage 8. Estimated revenue growth 9.
Net profit margin 10. Median recommendation (across major brokerages)
11. Location of firm’s headquarters 12. Stock exchange on which the firm
is listed

# Use cluster analysis to explore and analyze the given dataset as follows:

**a.** Use only the numerical variables (1 to 9) to cluster the 21
firms. Justify the various choices made in conducting the cluster
analysis, such as weights for different variables, the specific
clustering algorithm(s) used, the number of clusters formed, and so on.

**b.**Interpret the clusters with respect to the numerical variables
used in forming the clusters.

**c.** Is there a pattern in the clusters with respect to the numerical
variables (10 to 12)? (those not used in forming the clusters)

**d.** Provide an appropriate name for each cluster using any or all of
the variables in the dataset.

```{r}
# Load the dataset
pharma <- read.csv("/Users/stephengombos/Documents/KSU MBA PROGRAM/Fall 2025 FUNDAMENTALS OF MACHINE LEARNING (BA-64060-002)/CSV Files/Pharmaceuticals.CSV")

summary(pharma)

# Set row names to be the company names for easier identification in plots
# We use pharma$Name instead of Symbol for clarity
rownames(pharma) <- pharma$Name
```

### Separate Numerical and Categorical Data

We must cluster using only the nine numerical variables (1-9). We will
separate our data into two dataframes 
one for clustering and one for later interpretation.

```{r}
# Select only the numerical variables (columns 3 through 11) for clustering
pharma_numeric <- pharma[, 3:11]

# Select the categorical variables (cols 10-12) and identifiers for interpretation
pharma_categorical <- pharma[, c(1, 2, 12, 13, 14)]
```

### Scale the Data

K-means clustering is distance-based, so it's highly sensitive to the
scale of the variables. Variables with large variances (like
`Market_Cap`, in billions) will dominate the clustering process over
variables with small variances (like `Beta`).

To ensure each variable contributes (roughly) equally, we must
standardize the data. We will use z-score scaling, which transforms each
variable to have a mean of 0 and a standard deviation of 1. This
addresses the "weights for different variables" justification required
by the assignment.

```{r}
# Scale the data using the scale() function (creates z-scores)
pharma_scaled <- scale(pharma_numeric)

# View the first few rows of the scaled data
head(pharma_scaled, 3)
```

Note the negative and positive values, centered around 0

### Visualize Pairwise Distances

Before clustering, we can visualize the dissimilarity matrix to see if
clusters are even present. Lighter squares indicate firms that are very
different; darker squares indicate firms that are very similar.

```{r}
# Compute distances on scaled data
distance <- get_dist(pharma_scaled)

# Visualize the distance matrix
fviz_dist(distance)
```

#Interpretation: The heat map shows several dark blocks (e.g., in the
bottom-left and top-right), suggesting that natural groupings of similar
firms exist in the data. This confirms that cluster analysis is an
appropriate technique.

------------------------------------------------------------------------

## 2. Part (a): Conducting the Cluster Analysis

Part (a) asks us to cluster the firms using the numerical variables and
justify our choices, including the algorithm, weights, and number of
clusters.

-   **Algorithm:** We chose k-Means as it is a robust, efficient, and
    widely-used partitioning algorithm, suitable for this dataset's
    size.
-   **Weights:** We "weighted" all variables equally by standardizing
    the data (z-scores), as explained in the preparation step.
-   **Number of Clusters (k):** We must find the optimal number of
    clusters. We will use two common methods: the Elbow (WSS) method and
    the Silhouette method.

### Determine Optimal k (Number of Clusters)

We set a seed to ensure our results are reproducible, as k-Means
involves random starting points.

```{r}
set.seed(123)
```

## a) Elbow Method (Within-Cluster Sum of Squares)

This method plots the total within-cluster sum of squares (WSS) for
different values of k. We look for an "elbow" in the plot, where adding
another cluster stops providing a significant reduction in WSS.

```{r}
# Use fviz_nbclust to compute and plot the WSS
fviz_nbclust(pharma_scaled, kmeans, method = "wss")
```

## Interpretation:

The plot shows a very sharp bend (an "elbow") at k = 4. After k = 4, the
line becomes much flatter, suggesting diminishing returns from adding
more clusters. This method strongly suggests k = 4 is optimal.

## b) Silhouette Method

This method measures how well-separated the clusters are. A high average
silhouette width indicates that firms are well-matched to their own
cluster and poorly-matched to neighboring clusters.

```{r}
# Use fviz_nbclust to compute and plot the average silhouette width
fviz_nbclust(pharma_scaled, kmeans, method = "silhouette")
```

**Interpretation:** The silhouette method shows the highest average
silhouette width at **k = 5**.

### Justification for Choosing k = 4

The Elbow method points to **k = 4** [cite: 24], while the Silhouette
method suggests k = 5. For this business context, a smaller number of
clusters is often more interpretable and actionable for an analyst[cite:
5, 6]. Since k = 4 has strong statistical support from the elbow plot
and represents a more parsimonious model, we will proceed with **k =
4**.

### Final k-Means Model (k = 4)

We now run the final `kmeans` algorithm with `centers = 4`. We use
`nstart = 25`, which runs the algorithm 25 times with different random
starting centroids and picks the best result, helping to avoid a poor
local optimum.

```{r}
# Run k-means with k = 4 and 25 random starts
k_final <- kmeans(pharma_scaled, centers = 4, nstart = 25)

# View the cluster centers (in standardized units)
kable(k_final$centers, caption = "Cluster Centers (Standardized Z-Scores)")
```

**Interpretation of Standardized Centers:** This table shows the
*average standardized value* for each variable in each cluster. \*
**Positive values** are *above* the dataset's average for that variable.
\* **Negative values** are *below* the dataset's average. \*
**Example:** Cluster 2 has a very high `Market_Cap` (2.19) and
`Net_Profit_Margin` (1.33). Cluster 3 has a very high `PE_Ratio` (2.05)
and `Leverage` (1.81).

```{r}
# Number of firms in each cluster
k_final$size
```

### Visualize the Final Clusters

We can visualize the clusters using the first two principal components
(which capture the most variance in the data).

```{r visualize-clusters}
fviz_cluster(k_final, data = pharma_scaled,
             repel = TRUE, # Avoid label overlapping,
             main = "k-Means Cluster Plot (k = 4)")
```

### K-Medians (Manhattan Distance) Check

This is an alternative distance check, `kcca` (k-medians) with Manhattan
distance. K-medians is less sensitive to outliers than k-means.

```{r k-medians}
#kmeans clustering, using manhattan distance
k_medians = kcca(pharma_scaled, k=4, kccaFamily("kmedians"))
k_medians
```

------------------------------------------------------------------------

## 3. Part (b): Interpret Clusters (Numerical Variables)

Part (b) asks us to interpret the clusters with respect to the numerical
variables. The standardized centers above are hard to read. A *much*
better way is to look at the **mean of the *original* unscaled data**
for each cluster.

```{r}
# Add the cluster assignments back to the original numeric data
pharma_numeric_clustered <- pharma_numeric %>%
  mutate(Cluster = k_final$cluster)

# Calculate the mean of each variable by cluster
cluster_summary_numeric <- pharma_numeric_clustered %>%
  group_by(Cluster) %>%
  summarise_all(mean) %>%
  mutate(Cluster_Size = k_final$size) # Add cluster size for context

# Print a clean table
kable(cluster_summary_numeric, digits = 2, 
      caption = "Cluster Means (Original Unscaled Data)")
```

### Analysis of Numerical Profiles:

-   **Cluster 1 (Size = 8):** These firms have **average to low
    financial metrics** across the board. They have the *lowest*
    `Market_Cap` (\$43.7B), `ROE` (19.8%), and `ROA` (9.5%). Their
    `PE_Ratio` (22.3) and `Net_Profit_Margin` (15.5%) are also modest.
-   **Cluster 2 (Size = 4):** This is the **"Blue Chip / Market
    Leader"** group. These 4 firms have the *highest* `Market_Cap` (by
    far, \$132.0B), `ROE` (46.8%), `ROA` (17.8%), `Asset_Turnover`
    (0.95), and `Net_Profit_Margin` (22.4%). They also have very low
    `Leverage` (0.33).
-   **Cluster 3 (Size = 3):** This is a small, high-risk, high-growth
    group. They have a very high `Rev_Growth` (23.4%) and the *highest*
    `Leverage` (1.85). They also have the *highest* `PE_Ratio` (55.9,
    suggesting high market expectations) but the *lowest*
    `Net_Profit_Margin` (8.8%), indicating they are not yet highly
    profitable.
-   **Cluster 4 (Size = 6):** This group shows **strong profitability
    but negative growth**. They have solid `ROE` (23.3%) and `ROA`
    (11.7%), but they also have the *lowest* `Beta` (0.3), `PE_Ratio`
    (20.5), and `Rev_Growth` (a negative -1.4%). This suggests they are
    stable, mature companies, but their revenue is shrinking.

------------------------------------------------------------------------

## 4. Part (c): Cluster Patterns (Categorical Variables)

Part (c) asks if there is a pattern in the clusters with respect to the
categorical variables (10-12) that were *not* used in the analysis.

```{r}
# Add cluster assignments to the full, original dataset
pharma_clustered <- pharma %>%
  mutate(Cluster = k_final$cluster)
```

### Pattern by Median Recommendation

```{r}
rec_table <- table(pharma_clustered$Cluster, pharma_clustered$Median_Recommendation)
kable(rec_table, caption = "Cluster vs. Median Recommendation")
```

**Pattern:** Yes, there is a clear pattern.

-   **Cluster 1 (Low/Average):** This group is mixed, but it's the only
    one with "Moderate Sell" recommendations.

-   **Cluster 2 (Market Leaders):** 75% (3 of 4) have a "Moderate Buy"
    recommendation.

-   **Cluster 3 (High-Leverage/Growth):** 67% (2 of 3) have a "Moderate
    Buy" recommendation.

-   **Cluster 4 (Profitable/Shrinking):** 83% (5 of 6) have a "Hold"
    recommendation, which aligns with their negative revenue growth.

### Pattern by Location

```{r}
loc_table <- table(pharma_clustered$Cluster, pharma_clustered$Location)
kable(loc_table, caption = "Cluster vs. Location")
```

**Pattern:** Yes, a very strong pattern.

-   **Clusters 1, 2, and 3** are all **100% US-based firms**.

-   **Cluster 4** is **100% non-US firms** (UK, SWITZERLAND, GERMANY,
    FRANCE). This is a major finding: the clustering algorithm separated
    the firms geographically based *only* on their financial data.

### Pattern by Exchange

```{r}
ex_table <- table(pharma_clustered$Cluster, pharma_clustered$Exchange)
kable(ex_table, caption = "Cluster vs. Exchange")
```

**Pattern:** Yes, this is also distinct.

-   **Clusters 1, 2, and 4** are almost exclusively listed on the
    **NYSE**.

-   **Cluster 3 (High-Leverage/Growth):** Is the most diverse, with
    firms on AMEX and NASDAQ, which are exchanges often associated with
    smaller or higher-growth companies.

------------------------------------------------------------------------

## 5. Part (d): Naming the Clusters

Part (d) asks us to provide an appropriate name for each cluster. Based
on the analysis from (b) and (c):

1.  **Cluster 1: "US Mid-Caps"**
    -   **Justification:** This is the largest group, composed entirely
        of US firms (from part c) with financial metrics that are
        broadly average or slightly below average for this dataset (from
        part b). They don't stand out as market leaders or as high-risk.
2.  **Cluster 2: "US Blue Chip Leaders"**
    -   **Justification:** This group contains the 4 firms with the
        highest market cap, profitability (ROE, ROA, Margin), and asset
        turnover (from part b). They are all US-based and listed on the
        NYSE, and analysts rate them as "Moderate Buys" (from part c).
3.  **Cluster 3: "High-Growth / High-Leverage US Firms"**
    -   **Justification:** This small group is defined by its extremely
        high leverage and high expected revenue growth (from part b).
        Their high P/E ratios suggest the market has high expectations.
        They are US-based but listed on more growth-oriented exchanges
        (AMEX, NASDAQ) (from part c).
4.  **Cluster 4: "Mature European Giants"**
    -   **Justification:** This cluster's defining feature is that it
        contains *all* the non-US firms (from part c). Their financial
        profile (from part b) shows maturity: solid profitability but
        low (or negative) revenue growth and low-risk betas. Analysts
        predominantly have "Hold" ratings on them (from part c).

------------------------------------------------------------------------

**Comparison:** The cluster sizes for k-medians (Sizes: 4, 3, 5, 9) are
different from k-means (Sizes: 8, 4, 3, 6). This shows that the
clustering solution is *moderately sensitive* to the algorithm and
distance metric used. However, the k-means solution, with its strong
separation of US/non-US firms, appears to have found a very clear and
interpretable structure in the data. We stand by the k-means solution.
